---
title: "Machine Learning Project"
author: "Mary Ann D"
date: "November 21, 2015"
output: html_document
---
##Executive Summary
The goal of the project is to predict the manner in which the participants in a fitness study performed their excercises.  The data is provided in 2 files: training set and test set.  The "classe" variable in the training set is the predictor.
The data for this project come from this source: http://groupware.les.inf.puc-rio.br/har.  

###Data Processing
Create the environment and load the training data set.
```{r }
library(knitr)
opts_chunk$set(fig.width=12,fig.height=6,cache=TRUE,cache.path="cache/",fig.path="Figs/")
library(caret)
library(AppliedPredictiveModeling)
library(randomForest)
### Read the training data
trnfileUrl<- "pml-training.csv"
traindat<- read.csv(trnfileUrl, header=TRUE,sep = ",",na.strings = c("NA", "#DIV/0!","","-"))
```
###Exploratory Data Analysis
After loading the training data set, review the data and transform and coerce the data as neccesary.  The Data is a large data set of a mix of values and will be subsetted into a data set that is usable for fitting a model.
```{r }
removedat = is.na(traindat)     # Determine which variables are NA
keepcol<- nrow(traindat)*0.75 
removecol<- which( colSums(removedat)> keepcol)  #Flag the columns that are more than 75% not applicable
alldat = traindat[, -removecol]  # Remove those columns
alldat <- alldat[, -(1:7)]    # Remove the first 7 columns
table(sapply(alldat[1,], class))  # coerce the data
```
###Build the Regression Models  
####Split the data 
Split the training data and use the smaller portion as the Cross Validation data. 
```{r }
set.seed(1000)
inTrain <- createDataPartition(y=alldat$classe, p=0.6, list=FALSE)
mytrain <- alldat[inTrain, ]
myCV <- alldat[-inTrain, ]
```
####Model Prediction
I used the random forest model since it the most widely used model for large data sets.
Random forests correct for decision trees' habit of overfitting to their training set per the documentation. 
```{r }
modFit<- randomForest(classe ~., data= mytrain)
```
To see how accurate the model is, predict the model on the Cross Validation data.
```{r }
predictrf<- predict(modFit, myCV, type="class")
confusionMatrix(predictrf, myCV$classe)
```
We see that this model is 99% accurate
and the OOS(Out of Sample) rate is about .01%
  
  
##Appendix
Apply the model to the Test data provided. Read and process the data in the same way that the training data was evaluated. Code provided
```{r }
TestfileUrl<- "pml-testing.csv"
testdat<- read.csv(TestfileUrl, header=TRUE,sep = ",",na.strings = c("NA", "#DIV/0!","","-"))
### Subset and transform the data
removedat = is.na(testdat)
keepcol<- nrow(testdat)*0.75
removecol<- which( colSums(removedat)> keepcol)
tstalldat = testdat[, -removecol]
tstalldat <- tstalldat[, -(1:7)]
table(sapply(tstalldat[1,], class))
### Run the model on the Test Data
predicttst<- predict(modFit, tstalldat, type="class")
pml_write_files = function(x){
   n = length(x)
   for(i in 1:n){
      filename = paste0("problem_id_",i,".txt")
      write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
   }
}
pml_write_files(predicttst)
```
